\section{Background}\label{Background-background}

\subsection{Denotational Semantics}\label{Background-denotational-semantics}

A conventional Scott--Strachey denotational semantics of a programming
language consists of definitions of \emph{abstract syntax},
\emph{semantic domains}, and \emph{semantic functions}. The abstract
syntax uses a \emph{context-free grammar} to define sets of abstract
syntax trees (ASTs) that model the compositional structure of programs
and their phrases. The semantic domains are defined by equations in
terms of \emph{domain constructors}, and can be mutually recursive. The
semantic functions are defined \emph{inductively}, also with mutual
recursion, and map ASTs to their \emph{denotations}, which are elements
of domains. The denotation of an AST models its contribution to the
semantics of complete programs, and is defined \emph{compositionally} in
terms of the denotations of its direct components. The AST arguments of
semantic functions are enclosed by \(\llbracket\dots\rrbracket\), to
avoid potential confusion between abstract syntax terms and semantic
notation.

Domains were originally required to be continuous lattices
\cite{Scott1970OMT,Scott1972CL,Stoy1977DSS,Tennent1976DSP}.
Various more general notions of domain have been suggested (see
\cite{Abramsky1995DT})
but for us it is only important that endomaps between domains have
fixpoints, recursive domain equations have solutions, and each domain
comes with a bottom element. The notation for domain constructors and
their accompanying operations in a conventional Scott--Strachey
semantics, summarised below, is independent of the choice of domains.

Let \(D, E\) be domains with elements \(\delta : D\), \(\epsilon : E\).

\begin{itemize}
\item
  Every domain \(D\) has a \emph{bottom} element \(\bot_D\) that
  represents absence of information (e.g., due to an error or
  nontermination of a computation), usually written just \(\bot\).
\item
  A function \(\phi : D \to E\) is (Scott-) \emph{continuous} if it is
  monotone and preserves limits of directed subsets
  of~\(D\)~\cite{Abramsky1995DT}.
  The \emph{function domain} \(F = D \to E\) consists of all continuous
  (total) functions from \(D\) to \(E\). The set of all functions from a
  set~\(A\) to a domain also forms a domain, ordered pointwise.
  Functions between domains are defined using abstraction
  \(\lambda \delta.\epsilon\), application \(\phi\,\delta\), the
  operation \(\AgdaRef{fix}\) that maps each endofunction
  \(\phi : D \to D\) to its least fixed point, and the operations
  associated with other domain constructors.
\item
  For any set \(A\) the \emph{flat domain} \(A_\bot\) is formed by
  adding \(\bot\) as a fresh element. Functions on sets are implicitly
  extended to (continuous) functions on flat domains, returning~\(\bot\)
  when any argument is~\(\bot\). Elements~\(\tau\) of the domain of
  \emph{truth-values}
  \(\textbf{T} = \{ \text{true}, \text{false} \}_\bot\) are used in
  \emph{conditionals} written \(\tau \to \delta_1, \delta_2\), where
  \(\text{true} \to \delta_1, \delta_2\) is \(\delta_1\),
  \(\text{false} \to \delta_1, \delta_2\) is \(\delta_2\), and
  \(\bot_{\textbf{T}} \to \delta_1, \delta_2\) is \(\bot_D\).
\item
  The \emph{separated sum domain} \(X = \ldots + Y + \ldots\) consists
  of injected elements written `\(\upsilon \textsf{ in } X\)' (where
  \(\upsilon : Y\) for some summand \(Y\)) together with \(\bot_X\). The
  \(\textbf{T}\)-valued operation \(\chi \mathbin{\textsf{E}} Y\)
  (written \(\chi \in Y\) in {[}(Scheme){]}) tests whether \(\chi : X\)
  is the injection of some \(\upsilon : Y\); if so, \(\chi \mid Y\)
  projects \(\chi\) to \(\upsilon\), otherwise to \(\bot_Y\).
\item
  The \emph{product domain} \(P = D \times E\) consists of pairs
  \(\langle \delta, \epsilon \rangle\) with
  \(\bot_P = \langle \bot_D, \bot_E \rangle\). When \(\pi : P\), the
  operations \(\pi \downarrow 1\) and \(\pi \downarrow 2\) select its
  components; similarly for domains of \(n\)-\emph{tuples}~\(D^n\) and
  finite \emph{sequences}~\(D^*\). Further operations on~\(D^*\) include
  the empty sequence~\(\langle \rangle\),
  concatenation~\(\delta^*_1 \mathbin{\S} \delta^*_2\),
  length~\(\text{\#}\,\delta^*\), \(n\)th~component
  \(\delta^* \downarrow n\), and \(n\)th~tail
  \(\delta^* \mathbin{\dagger} n\) (\(n \geq 1\)).
\end{itemize}

Robert Tennent's highly accessible tutorial on denotational
semantics~\cite{Tennent1976DSP}
includes illustrative examples using the above notation (partly in
continuation-passing style) and further details of the conventional
Scott--Strachey style.

\subsection{Agda}\label{Background-agda}

Agda
\cite{Agda-Language,Agda-Wikipedia}
is both a strongly typed functional \emph{programming language} with
support for first-class \emph{dependent types} and a \emph{proof
assistant} based on the Curry--Howard correspondence between
propositions and types. A number of features set Agda apart from a
typical functional programming language:

\begin{itemize}
\item
  Types in Agda are first-class values of a \emph{universe}
  \(\textsf{Set}_i\) where \(\textsf{Set} = \textsf{Set}_0\), and each
  universe \(\textsf{Set}_i\) belongs to the next universe
  \(\textsf{Set}_{i+1}\).
\item
  Agda has \emph{dependent function types} \AgdaRef{(x\ :\ A)\ →\ B} or
  \AgdaRef{∀\ x\ →\ B} where the type \AgdaRef{B} can depend on the value
  \AgdaRef{x}.
\item
  All functions in Agda are \emph{total}: evaluating a function call is
  guaranteed to return a value of the correct type in finite time. To
  ensure totality, Agda's type checker includes a termination check for
  recursive definitions, a positivity check for inductive datatypes, and
  a consistency check for universe levels.
\end{itemize}

Thanks to its dependent types and totality, Agda's type system can be
used as a higher-order logic for writing mathematical statements and
proofs, where types correspond to propositions and type checking
corresponds to checking the validity of the proof.

\subsubsection*{Notes on some of Agda's syntax and
features}\label{Background-notes-on-some-of-agdas-syntax-and-features}
\addcontentsline{toc}{subsubsection}{Notes on some of Agda's syntax and
features}

\begin{description}
\tightlist
\item[Comments.]
End-of-line comments are initiated by \AgdaRef{-\/-}, while multi-line
comments are between \AgdaRef{\{-} and \AgdaRef{-\}}.
\item[Naming.]
Declared symbols, variables, and type constructors all share a single
namespace. Names can be any sequence of non-whitespace ASCII and Unicode
characters except \AgdaRef{.;\{\}()@"}, excluding reserved symbols and
keywords such as \AgdaRef{→} or \AgdaRef{where}. Underscores in names play
a special role for mixfix notation.
\item[Mixfix notation.]
Agda supports \emph{mixfix notation} for defining operators, with
underscores in the name of a symbol indicating argument positions. For
example, if we declare \AgdaRef{\_+\_\ :\ Nat\ →\ Nat\ →\ Nat}, we can
use it as \AgdaRef{1\ +\ 1} (the spaces are required, since \AgdaRef{1+1}
is a valid Agda name!).
\item[Anonymous functions.]
Lambda abstractions use the syntax \AgdaRef{λ\ x\ →\ u} instead of
\AgdaRef{λ\ x.\ u}.
\item[Implicit arguments.]
Arguments marked by single curly braces \AgdaRef{\{…\}} in the type of a
symbol are considered to be \emph{implicit}. These arguments may be
omitted, and are then inferred by Agda's type checker.
\item[Type classes.]
Agda has no direct support for type classes, but they can be simulated
using Agda's \emph{instance arguments} to resolve type class instances.
Instance arguments are marked by double curly braces \AgdaRef{\{\{…\}\}}
and are resolved automatically by using definitions marked as
\AgdaRef{instance}.
\item[Rewrite rules.]
Agda has support for \emph{rewrite rules}
\cite{cockx:LIPIcs.TYPES.2019.2}
that are applied automatically during type checking. Rewrite rules are
declared by marking an equality proof \AgdaRef{eq\ :\ a\ ≡\ b} with a
\AgdaRef{\{-\#\ REWRITE\ eq\ \#-\}} pragma. Agda can optionally check
confluence of rewrite rules, but currently does not check their
termination. Since only \emph{proven} (or postulated) equalities can be
added as rewrite rules, non-confluence and non-termination cannot affect
the soundness of Agda's type checker, only its
completeness~\cite{10.1145/3434341}.
\end{description}
